Astr 513: Final Project

### 2016-05-02 Final_Project/ms.tex
Expanded on the correlation matrices. Added some comments.

\section{Type \rom{1}a Supernovae}
\label{sec:SNIa}
\EW{Should I cite the Nobel Prize information document where I found out about 
the stuff in the last few sentences? I'm not entirely sure what that would be considered...}
A type \rom{1}a supernova (\sna{}) is a supernova with very little hydrogen or
helium absorption. It is believed that the progenitor for these special
supernovae are white dwarfs that are part of a binary system. These white
dwarfs accrete mass from their companion stars or merge with a companion white
dwarf or neutron star, causing them to exceed the Chandrasekhar limit
($1.44 M_\odot$). When this happens, carbon and oxygen fusion begins in the
core, and the temperature increases. However, the white dwarf does not expand
and cool as a main sequence star would, since it is supported by degeneracy
pressure rather than thermal, so the core experiences a runaway fusion
reaction, releasing $\sim 10^{44}$ J of energy \citep[see, e.g.,][chap. 18.4]{ryden2010}, which is enough to overcome the gravitational energy and the star explodes. Because all \snea{} are caused by a similar reaction, they all have similar luminosities, and more importantly, the light curves all have the same shape: a sharp maximum followed by a smooth, gradual decay. Studying nearby \snea{} has shown that the rate of decline of the light curve is correlated to the peak luminosity, and this correlation has been calibrated using Cephied variable stars located in the same galaxies as nearby \sna{}. With this information, we can estimate peak luminosity of an \sna{} from its light curve, thereby providing us with a standard candle with which to measure distance. They were first used in this way in the 1990s by two independent teams, the Supernova Cosmology Project and the Cal\'{a}n/Tololo Supernova Survey. It was using distances from \snea{} from these two projects that \citet{riess1998} and \citet{perlmutter1999} were able to discover the accelerating expansion rate of the Universe in 1998. It was for this work that Saul Perlmutter, Brian P. Schmidt, and Adam G. Reiss won the Nobel Prize in physics in 2011.

@@

\subsection{The JLA sample}
\label{sec:betoule}

\citet{betoule2014} define their ``joint light-curve analysis'' (JLA) 
sample, a total of 740 \snea{} compiled from a combination of 
data from the Sloan Digital Sky 
Survey (SDSS), the Supernova Legacy Survey (SNLS), the Hubble Space 
Telescope (HST), and several other experiments. The 613 SDSS \snea{} 
come from the SDSS supernova survey (SDSS-SNS) of SDSS-II \citep{sako2014} 
with redshifts of $0.05 < z < 0.4$, and 374 of them are selected from those 
that were confirmed in the spectroscopic follow-up. The rest of the 
\sne{} come from the compilation by \citet{conley2011}, which includes 
242 spectroscopically confirmed \snea{} from SNLS at redshifts $0.2 < z < 1$, 
14 \snea{} from HST at very high redshifts of $0.7 < z < 1.4$, and 123 
low redshift ($z < 0.08$) from a variety of sources. We made use of the 
publicly available distance moduli, redshifts, and covariance 
matrix provided by \citet{betoule2014}\footnote{Data is available 
for download via 
\url{http://cdsarc.u-strasbg.fr/viz-bin/qcat?J/A+A/568/A22}}. 
We note that the distance moduli are binned, so that they are only computed and 
given on a grid of redshifts. The distance modulus can then be well 
approximated by a piecewise linear function of $\log(z)$ that is given 
in a single redshift bin $z_b \le z < z_{b+1}$ as 
\citep[see][Appendix E.1]{betoule2014}
\begin{equation}
\label{eq:binnedMu}
\bar{\mu}(z) = (1 - \alpha) \mu_b + \alpha \mu_{b+1},
\end{equation}
where they have defined 
$\alpha = \log\left(\frac{z}{z_b}\right)/\log\left(\frac{z_{b+1}}{z_b}\right)$ 
and $\mu_i$ is the distance modulus at control point $z_i$. 
Therefore, the data and covariance matrix are merely given at the 
31 control points log-spaced in the redshift range $0.01 < z < 1.3$.

\subsection{The Pan-STARRS1 sample}
\label{sec:rest}

\citet{rest2014} uses the first 1.5 year of the first telescope of 
the Panoramic Survey Telescope and Rapid Response System (Pan-STARRS1) 
and a compilation of low-redshift {\sna} studies. The low-redshift 
sample consists of {\snea} collected by Calan/Tololo (Hamuy et al. 1996a), 
CfA1 (Riess et al. 1999), CfA2 (Jha et al. 2006), CfA3 (Hicken et al. 2009a), 
CSP (Contreras et al. 2010), and CfA4 (Hicken et al. 2012). After 
three quality cuts, 113 spectroscopically confirmed {\snea} from 
Pan-STARRS1 ($0.03 < z < 0.65$) and 222 light curves from 197 
low-redshift {\snea} are used for cosmological analysis.

The publicly available data from \citet{rest2014} contains all light 
curves, light curve fit information, covariance matrix, and each {\sn} 
host information.\footnote{Data can be accessed at 
\url{http://ipp0012.ifa.hawaii.edu/SN1aLC/}}. For our analysis, we 
utilized the redshifts, distance moduli, and covariance matrix of the 
final sample of 310 {{\snea} or 335 light curves.

\subsection{Covariances}
\label{sec:covariances}
\LC{Wrote about Rest's cov matrix. Still need description of Betoule's.}
\EW{Wrote about Beoule's covariance matrix.}

\begin{figure}
 \includegraphics[width=\linewidth]{figures/Betoule_correlation.eps}
 \caption{Correlation matrix of $\mu_0(z)$ from the
 \citet{betoule2014} sample.}
 \label{fig:cmatrix}
\end{figure}

The publicly released covariance matrix from \citet{rest2014} 
is comprised of systematic errors for the final sample. These systematics, 
defined in \citet{scolnic2014}, include flux calibration errors, 
selection effects, light curve fitting, host galaxy relations, Milky 
Way extinction corrections, and coherent velocity flows. Calculating 
the covariance matrix involves varying a systematic parameter and determining 
the difference between the original distance and the new distance. 
The covariance between {\sn} $i$ and {\sn} $j$ is 
\begin{equation}
\label{eq:RestCovMatrix}
C_{ij}=\displaystyle \sum_{k = 1}^K \left(\frac{\partial \mu_i}{\partial S_k}\right)\left(\frac{\partial \mu_j}{\partial S_k}\right)\left(\sigma_{S_k}\right)^2,
\end{equation}
where the summation is over $K$ systematics, $\sigma_{S_k}$ is each 
systematic error magnitude, and $\partial \mu$ is the difference in 
distance modulus from varying a systematic parameter.

@@

\begin{figure}
 \includegraphics[width=\linewidth]{figures/Rest_correlation.eps}
 \caption{Correlation matrix of $\mu_0(z)$ from the
 \citet{rest2014} sample.}
 \label{fig:cmatrix}
\end{figure}

>>>>

\section{Type \rom{1}a Supernovae}
\label{sec:SNIa}
\EW{Should I cite the Nobel Prize information document where I found out about 
the stuff in the last few sentences? I'm not entirely sure what that would be considered...}
\TM{No, because it is common knowledge. I think though that since this is meant to emulate
  a research paper that we don't need to point out that it was what the NP was awarded for.}
A type \rom{1}a supernova (\sna{}) is a supernova with very little hydrogen or
helium absorption. It is believed that the progenitor for these special
supernovae are white dwarfs that are part of a binary system. These white
dwarfs accrete mass from their companion stars or merge with a companion white
dwarf or neutron star, causing them to exceed the Chandrasekhar limit
($1.44 M_\odot$). When this happens, carbon and oxygen fusion begins in the
core, and the temperature increases. However, the white dwarf does not expand
and cool as a main sequence star would, since it is supported by degeneracy
pressure rather than thermal, so the core experiences a runaway fusion
reaction, releasing $\sim 10^{44}$ J of energy \citep[see, e.g.,][chap. 18.4]{ryden2010}, 
which is enough to overcome the gravitational energy and the star explodes. Because all 
\snea{} are caused by a similar reaction, they all have similar luminosities, and more 
importantly, the light curves all have the same shape: a sharp maximum followed by a smooth, 
gradual decay. Studying nearby \snea{} has shown that the rate of decline of the light curve 
is correlated to the peak luminosity, and this correlation has been calibrated using Cephied 
variable stars located in the same galaxies as nearby \sna{}. With this information, one can 
estimate peak luminosity of an \sna{} from its light curve, thereby providing a 
standard candle with which to measure distance. They were first used in this way in the 
1990s by two independent teams, the Supernova Cosmology Project and the Cal\'{a}n/Tololo 
Supernova Survey. It was using distances from \snea{} from these two projects that 
\citet{riess1998} and \citet{perlmutter1999} were able to discover the accelerating expansion 
rate of the Universe in 1998. 
%It was for this work that Saul Perlmutter, Brian P. Schmidt, 
%and Adam G. Reiss won the Nobel Prize in physics in 2011.
\TM{Changed grammer in this section, since ``we'' did not actually do these things.
  Review and collapse bubbles.}

@@

\subsection{The JLA sample}
\label{sec:betoule}
\citet{betoule2014} define their ``joint light-curve analysis'' (JLA) 
sample, a total of 740 \snea{} compiled from a combination of 
data from the Sloan Digital Sky 
Survey (SDSS), the Supernova Legacy Survey (SNLS), the Hubble Space 
Telescope (HST), and several other experiments. The 613 SDSS \snea{} 
come from the SDSS supernova survey (SDSS-SNS) of SDSS-II \citep{sako2014} 
with redshifts of $0.05 < z < 0.4$, and 374 of them are selected from those 
that were confirmed in the spectroscopic follow-up. The rest of the 
\sne{} come from the compilation by \citet{conley2011}, which includes 
242 spectroscopically confirmed \snea{} from SNLS at redshifts $0.2 < z < 1$, 
14 \snea{} from HST at very high redshifts of $0.7 < z < 1.4$, and 123 
low redshift ($z < 0.08$) from a variety of sources. We made use of the 
publicly available distance moduli, redshifts, and covariance 
matrix provided by \citet{betoule2014}\footnote{Data is available 
for download via 
\url{http://cdsarc.u-strasbg.fr/viz-bin/qcat?J/A+A/568/A22}}. 
We note that the distance moduli are binned, so that they are only computed and 
given on a grid of redshifts. The distance modulus can then be well 
approximated by a piecewise linear function of $\log(z)$ that is given 
in a single redshift bin $z_b \le z < z_{b+1}$ as 
\citep[see][Appendix E.1]{betoule2014}
\begin{equation}
\label{eq:binnedMu}
\bar{\mu}(z) = (1 - \alpha) \mu_b + \alpha \mu_{b+1},
\end{equation}
where they have defined 
$\alpha = \log\left(\frac{z}{z_b}\right)/\log\left(\frac{z_{b+1}}{z_b}\right)$ 
and $\mu_i$ is the distance modulus at control point $z_i$. 
Therefore, the data and covariance matrix are merely given at the 
31 control points log-spaced in the redshift range $0.01 < z < 1.3$.

\subsection{The Pan-STARRS1 sample}
\label{sec:rest}
\citet{rest2014} uses the first 1.5 year of the first telescope of 
the Panoramic Survey Telescope and Rapid Response System (Pan-STARRS1) 
and a compilation of low-redshift {\sna} studies. The low-redshift 
sample consists of {\snea} collected by Calan/Tololo (Hamuy et al. 1996a), 
CfA1 (Riess et al. 1999), CfA2 (Jha et al. 2006), CfA3 (Hicken et al. 2009a), 
CSP (Contreras et al. 2010), and CfA4 (Hicken et al. 2012). After 
three quality cuts, 113 spectroscopically confirmed {\snea} from 
Pan-STARRS1 ($0.03 < z < 0.65$) and 222 light curves from 197 
low-redshift {\snea} are used for cosmological analysis.

The publicly available data from \citet{rest2014} contains all light 
curves, light curve fit information, covariance matrix, and each {\sn} 
host information.\footnote{Data can be accessed at 
\url{http://ipp0012.ifa.hawaii.edu/SN1aLC/}}. For our analysis, we 
utilized the redshifts, distance moduli, and covariance matrix of the 
final sample of 310 {{\snea} or 335 light curves.
\TM{There is some kind of undefined reference error in the above line. 
  It doesn't alter the pdf document, but can we figure out what is going on?}

\subsection{Covariances}
\label{sec:covariances}
The publicly released covariance matrix from \citet{rest2014} 
is comprised of systematic errors for the final sample. These systematics, 
defined in \citet{scolnic2014}, include flux calibration errors, 
selection effects, light curve fitting, host galaxy relations, Milky 
Way extinction corrections, and coherent velocity flows. Calculating 
the covariance matrix involves varying a systematic parameter and determining 
the difference between the original distance and the new distance. 
The covariance between {\sn} $i$ and {\sn} $j$ is 
\begin{equation}
\label{eq:RestCovMatrix}
C_{ij}=\displaystyle \sum_{k = 1}^K \left(\frac{\partial \mu_i}{\partial S_k}\right)\left(\frac{\partial \mu_j}{\partial S_k}\right)\left(\sigma_{S_k}\right)^2,
\end{equation}
where the summation is over $K$ systematics, $\sigma_{S_k}$ is each 
systematic error magnitude, and $\partial \mu$ is the difference in 
distance modulus from varying a systematic parameter.

@@

For both of these data sets, the covariance matrix can be visualized using a related
quantity known as the correlation matrix $Corr$. When two quantities are (anti)correlated,
the value of one quantity (decreases)increases when the other quantity increases.
If the variables one has measured are independent, then the correlation between
differen variables should be $\approx0$.

A correlation matrix for $N$ variables is $N\times N$ in size, and the $ij$th
element is the correlation between the $i$th and $j$th variables. Therefore, 
the correlation matrix for \citet{betoule2014} is $31\times31$, while
for \citet{rest2014} is $335\times335$.

The correlation matrix is calculated from the covariance matrix according to
\begin{equation}
  \label{eq:correlation}
  Corr_{i,j} = \frac{\mathbf{C}_{i,j}}{\sqrt{\mathbf{C}_{i,i}}\sqrt{\mathbf{C}_{j,j}}}.
\end{equation}
Note that the correlation of a variable with itself is 1 by definition.
The correlation matrices for both of our data sets can be seen below in
figures \autoref{fig:cmatrix_betoule} and \autoref{cmatrix_rest}.

\TM{TODO: Discuss how the correlation matrices for the two data sets are different
  and hypothesize on what this might mean.}

\begin{figure}
 \includegraphics[width=\linewidth]{figures/Betoule_correlation.eps}
 \caption{Correlation matrix of $\mu_0(z)$ from the
 \citet{betoule2014} sample.
 \TM{Need to improve this caption.}}
 \label{fig:cmatrix_betoule}
\end{figure}

\begin{figure}
 \includegraphics[width=\linewidth]{figures/Rest_correlation.eps}
 \caption{Correlation matrix of $\mu_0(z)$ from the
 \citet{rest2014} sample.
 \TM{Need to improve this caption.}}
 \label{fig:cmatrix_rest}
\end{figure}
